---
title: "An Introduction to Databases"
author: "Zach"
date: "2020-04-13"
output: html_document
---
    
    
```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE)
```

## Databases - Storing and Managing Data
This week as part of the series, Kyle introduced us to some basic database management
by moving the csv data that we dowloaded from the [*JHU CSSE database*](https://github.com/CSSEGISandData/COVID-19) to a local Postgres instance on his machine and demoing some simple commands.


### Some basic commands
Here is a table of some basic SQL commands:
```{r echo=FALSE}
library(data.table)
library(knitr)
library(magrittr)
defs.list <- list('Fetch data from the database (i.e. query the database)',
                  'Edit existing rows in a table',
                  'Add new rows to a table',
                  'Delete rows from a table'
)

defs.dt <- data.table('Command'=list('SELECT','UPDATE','INSERT','DELETE'),
                      'Definition'=defs.list)
kable(defs.dt)
```


    

## Further reading
See the following links for information on the following topics:

* [*Codecademy article on databases*](https://www.codecademy.com/articles/what-is-rdbms-sql).
Codecademy is also a great place for you to get started with a number of programming languages
on your own with simple introductory lessons for FREE.

