---
title: "Data Download and Sanitation (and our first plot!)"
author: "Zach"
date: "2020-03-30"
output: html_document
---


```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE)
```


## Libraries
At the top of any script we write, first we include all the libraries used in this
analysis.   These "libraries" are packages of code that we import to our script 
so we may use some of their handy features.  For example, the **xml2** library allows
us to download a webpage and extract all of the text from it in just a few simple
lines of code!  A few commonly used libraries include **data.table**, **dplyr**, and 
**magrittr**, all of which we will go into later.
```{r libraries}
library(magrittr)
library(xml2)
library(lubridate)
library(data.table)
library(dplyr)
library(ggplot2)
```


## Set Country Name
Setting this line allows us to easily switch between countries.
```{r country-switch}
# Choose country to look at
country.switch <- 'France'
```

## Set paths
This line is purely procedural.  Here we tell R in a platform-agnostic way where 
to put on your computer the downloaded csv data as well as the R scripts.  In 
other words, R makes a guess at where your home directory is, regardless of whether
you are using Windows, MacOS, or Linux.
```{r set-paths}
git.path <- Sys.getenv('HOME')  # Where the base COVID19-Data-Exploration folder lives.
```


## Fetching the CSV data from Github
Here's where we start to get our hands dirty.  In the 3 code chunks below, we 
first download the Github webpage that has all of the csv files in it and extract 
the text (*scrape-webpage*).  Second, we look through all of that text for any
string of characters in the date format YYYY-MM-DD, since that is the naming 
convention for all of the csv files, and make a list of those dates (*parse-dates*).  
Finally, we use the list of dates (corresponding to the names of the csv files)
to download the raw comma-separated data and save it locally to our computer 
(*fetch-data*).

```{r scrape-webpage}
# Pull in list of daily data. 
xml.path <- 'https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports'
report.file <- download_html(xml.path)
html.read <- read_html(report.file)
csv.list <- xml_text(html.read) %>% strsplit(split='\\n') %>% unlist
```

```{r parse-dates}
parse_dates.fn <- function(csv.list) {
    # Parse html for the csv names
    dates.list <- lapply(csv.list, function(x) {
    if (grepl(pattern='.csv', x, fixed=TRUE)) { 
        regex <- regexpr('\\d{2}-\\d{2}-\\d{4}.csv', x)
        return(substr(x, start=regex[[1]], stop=regex[[1]]+attr(regex, 'match.length')))
        }
    })
    
    return(dates.list)
}

dates.list <- parse_dates.fn(csv.list)
dates.list[sapply(dates.list, is.null)] <- NULL  # Drop null values in list
curr_csvs.list <- list.files(paste0(git.path, '/Code/COVID19-Data-Exploration/data/'))
```

```{r fetch-data}
# Fetch raw csv data
raw.path <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/'
suppress <- lapply(dates.list, function(x, raw.path) {
    if (!(x %in% curr_csvs.list))  # Only download most recent csv
        download.file(paste0(raw.path, x), paste0(git.path,'/Code/COVID19-Data-Exploration/data/',x))
}, raw.path)
```


## Data Sanitization -- Making our Data Tidy
Now that we have our data downloaded, we are ready to do some analysis, right?  
Yes, but only if our data are in the right format.  In R, making one's data "tidy"
doesn't just mean it is clean and neat in the general sense of those words; rather, 
it means formatting one's data in a way that makes it easy to split it up into 
small chunks, apply a set of operations to each chunk, and then recombine the 
chunks back into a new dataset.  This split-apply-combine approach is a common
motif in data science and, once you get the hang of it, is a quite intuitive and
useful way to handle your data!

```{r trim-func}
# trim whitespace and any trailing digits
trim <- function(x) return(tstrsplit(x, "\\s|[A-Z]", keep=1) %>%unlist)
```

```{r tidy-up}
data.dt <- lapply(dates.list, function(x) {  # Split
    
    # -- Apply -- #
    # Fix column names and standardize date formats
    tmp.dt <- fread(paste0(git.path,'/Code/COVID19-Data-Exploration/data/',x)) %>% data.table
    colnames(tmp.dt) <- colnames(tmp.dt) %>% gsub('[ \\/]', '_', .)  # sub spaces and slashes for underscore
    # Fix date formatting for first 10 days
    tryCatch({
        tmp.dt$Last_Update <- tmp.dt$Last_Update %>% 
            trim %>% 
            parse_date_time(orders=c('%m/%d/%y','%m/%d/%Y','%Y-%m-%d'))
        return(tmp.dt)},
        warning = function(w) {
            message(paste('Warning!  Check file:', x))
        },
        error = function(e) {
            message(paste('Error!  Check file', x))
        }
    )
    # -- End of Apply -- #
    
}) %>% rbindlist(fill=TRUE) # Combine
```


## Filtering the Dataset
```{r filter-by-US}
data.dt[data.dt$Recovered%>%is.na, 'Recovered'] <- 0
data.dt[data.dt$Deaths%>%is.na, 'Deaths'] <- 0
data.dt[data.dt$Confirmed%>%is.na, 'Confirmed'] <- 0

country_cases.dt <- data.dt[data.dt$Country_Region==country.switch,] %>% 
    group_by(Last_Update) %>% 
    summarize(Recovered=max(Recovered),Deaths=max(Deaths)) %>% 
    data.table

# Make separate data table for confirmed, since it can be an order of magnitude
# higher than recovered/death reports.
country_confirmed.dt <- data.dt[data.dt$Country_Region==country.switch,] %>% 
    group_by(Last_Update) %>% 
    summarize(Confirmed=max(Confirmed)) %>% data.table
```


## Our First Plot
```{r plot-cases}
melted.dt <- melt(country_confirmed.dt, id.vars='Last_Update', variable.name = 'Cases', 
                  value.name='Number.Reported')
ggplot(melted.dt, aes(x=Last_Update, y=Number.Reported)) +
    geom_point() + 
    geom_line()


melted.dt <- melt(country_cases.dt, id.vars='Last_Update', variable.name = 'Cases', 
                  value.name='Number.Reported')
ggplot(melted.dt, aes(x=Last_Update, y=Number.Reported, color=Cases)) +
    geom_point() + 
    geom_line()
```

    

