---
title: "Data Download and Sanitation (and our first plot!)"
author: "Zach"
date: "2020-03-30"
output: html_document
---


```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE)
```

# March 30, 2020
Hello, and welcome to a summary of the first training session for bio-group technical
development!  This is an Rmarkdown document that will walk you through what we 
did in the first session on March 30, 2020.  Here we will download the data from the
[*JHU CSSE database*](https://github.com/CSSEGISandData/COVID-19) and plot the
reported confirmed cases, deaths, and recoveries since January 22, 2020.


## Libraries
At the top of any script we write, first we include all the libraries used in this
analysis.   These "libraries" are packages of code that we import to our script 
so we may use some of their handy features.  For example, the **xml2** library allows
us to download a webpage and extract all of the text from it in just a few simple
lines of code!  A few commonly used libraries include **data.table**, **dplyr**, and 
**magrittr**, all of which we will go into later.
```{r libraries}
library(magrittr)
library(xml2)
library(lubridate)
library(data.table)
library(dplyr)
library(ggplot2)
```


## Set paths
This line is purely procedural.  Here we tell R in a platform-agnostic way where 
to put on your computer the downloaded csv data as well as the R scripts.  In 
other words, R makes a guess at where your home directory is, regardless of whether
you are using Windows, MacOS, or Linux.
```{r set-paths}
git.path <- Sys.getenv('HOME')  # Where the base COVID19-Data-Exploration folder lives.
```


## Fetching the CSV data from Github
Here's where we start to get our hands dirty.  In the 3 code chunks below, we 
first download the Github webpage that has all of the csv files in it and extract 
the text  Second, we look through all of that text for any
string of characters in the date format YYYY-MM-DD, since that is the naming 
convention for all of the csv files, and make a list of those dates.  
Finally, we use the list of dates (corresponding to the names of the csv files)
to download the raw comma-separated data and save it locally to our computer.

```{r scrape-webpage}
# Pull in list of daily data. 
xml.path <- 'https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports'
report.file <- download_html(xml.path)
html.read <- read_html(report.file)
csv.list <- xml_text(html.read) %>% strsplit(split='\\n') %>% unlist
```

```{r parse-dates}
parse_dates.fn <- function(csv.list) {
    # Parse html for the csv names
    dates.list <- lapply(csv.list, function(x) {
    if (grepl(pattern='.csv', x, fixed=TRUE)) { 
        regex <- regexpr('\\d{2}-\\d{2}-\\d{4}.csv', x)
        return(substr(x, start=regex[[1]], stop=regex[[1]]+attr(regex, 'match.length')))
        }
    })
    
    return(dates.list)
}

dates.list <- parse_dates.fn(csv.list)
dates.list[sapply(dates.list, is.null)] <- NULL  # Drop null values in list
curr_csvs.list <- list.files(paste0(git.path, '/Code/COVID19-Data-Exploration/data/'))
```

```{r fetch-data}
# Fetch raw csv data
raw.path <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/'
suppress <- lapply(dates.list, function(x, raw.path) {
    if (!(x %in% curr_csvs.list))  # Only download most recent csv
        download.file(paste0(raw.path, x), paste0(git.path,'/Code/COVID19-Data-Exploration/data/',x))
}, raw.path)
```


## Data Sanitization -- Making our Data Tidy
Now that we have our data downloaded, we are ready to do some analysis, right?  
Yes, but only if our data are in the right format.  In R, making one's data "tidy"
doesn't just mean it is clean and neat in the general sense of those words; rather, 
it means formatting one's data in a way that makes it easy to split it up into 
small chunks, apply a set of operations to each chunk, and then recombine the 
chunks back into a new dataset.  This split-apply-combine approach is a common
motif in data science and, once you get the hang of it, is a quite intuitive and
useful way to handle your data!

```{r trim-func}
# trim whitespace and any trailing digits.  Used to set up the file timestamps
# to easily organize the data by day.
trim <- function(x) return(tstrsplit(x, "\\s|[A-Z]", keep=1) %>%unlist)
```

```{r tidy-up}
data.dt <- lapply(dates.list, function(x) {  # SPLIT  up dates.list into chunks.
    
    # -- APPLY -- #
    # Fix column names and standardize date formats
    tmp.dt <- fread(paste0(git.path,'/Code/COVID19-Data-Exploration/data/',x)) %>% data.table
    colnames(tmp.dt) <- colnames(tmp.dt) %>% gsub('[ \\/]', '_', .)  # sub spaces and slashes for underscore
    # Fix date formatting for first 10 days
    tryCatch({
        tmp.dt$Last_Update <- tmp.dt$Last_Update %>% 
            trim %>% 
            parse_date_time(orders=c('%m/%d/%y','%m/%d/%Y','%Y-%m-%d'))
        return(tmp.dt)},
        warning = function(w) {
            message(paste('Warning!  Check file:', x))
        },
        error = function(e) {
            message(paste('Error!  Check file', x))
        }
    )
    # -- End of APPLY -- #
    
}) %>% rbindlist(fill=TRUE) # COMBINE all the file chunks into a single data table.
```


## Filtering the Dataset
Now that we have the dataset downloaded and organized in a way that is easy to work with 
(i.e. a single combined data table), we can now think of questions that we want to
answer using the data.  This is an early step in the exploratory data analysis phase.
A first probing question may be, "For our country of interest, what are the number 
of confirmed cases, reported deaths, and reported recoveries?"

To answer this question, we are going to make a few simple plots to visually see
how the number of confirmed COVID-19 cases, the number of reported deaths due to
the virus, and the number of reported cases have changed over time.

### Replace NA's with zeros
It is often convenient to replace NA values in a dataset with zeros, since NA 
values in R (as well as other programming languages) can have weird or behavior 
that may be different from what you are expecting (e.g. summing values in
column that has an NA in it will return NA).
```{r replace-NAs}
data.dt[data.dt$Confirmed%>%is.na, 'Confirmed'] <- 0
data.dt[data.dt$Deaths%>%is.na, 'Deaths'] <- 0
data.dt[data.dt$Recovered%>%is.na, 'Recovered'] <- 0

data.dt
```

### Select Country
Choose country to look at.  We are using Italy here because it is more easy
to verify our results with those reported by the CSSE data dashboard.
```{r country-switch}
country.switch <- 'Italy'
```

### Filter by country, group by day, and summarize
This is our first step where we really start to manipulate the data to answer our
question of interest.  All of the organizing we did to our data before this was so
that we can use these powerful filtering, grouping and summarizing tools to select
the chunks of our data that we are interested in. 

These 4 lines make heavy use of %>%', which is the piping operator from the 
**magrittr** library, which conveniently allows us to string together multiple
operations in a single step.  In the 4 lines below we:

1. SELECT from the dataset only the rows that match our country of interest.
2. GROUP the selected rows by their date (to get a chronological timeline of
   the number of cases in each category).
3. SUMMARIZE all the observations (i.e. rows) for each date into a single value.
       We take only the maximum value reported for each day, since that will reflect
       the number of reported confirmed cases/deaths/recoveries for that day.
       
```{r filter-group-summarize}
country_cases.dt <- data.dt[data.dt$Country_Region==country.switch,] %>%  # SELECT country of interest
    group_by(Last_Update) %>%   # GROUP by date
    summarize(Recovered=max(Recovered),Deaths=max(Deaths)) %>% # SUMMARIZE by max daily value 
    data.table  # Ensure the data table remains a data table
country_cases.dt
```

Here, we make separate data table for confirmed cases, since they are up to an 
order of magnitude higher than recovered/death reports.  Visually it is easier to 
distinguish the plots if they are separated, since if all 3 were on the same plot
the plots for recovered/deaths would be very small compared to the number of confirmed
cases.  The steps are the same we just did for the previous data.table
```{r filter-group-summarize-confirmed}
country_confirmed.dt <- data.dt[data.dt$Country_Region==country.switch,] %>% 
    group_by(Last_Update) %>% 
    summarize(Confirmed=max(Confirmed)) %>% 
    data.table
```


## Our First Plots
Finally, let us plot the data!  Here we make use of the **ggplot2** library to
plot our data.  In this first plot, we show the number of reported cases (y-axis)
over time (x-axis).  Again, the plot is showing the maximum number of reported
cases each day since Jan. 22, 2020 for our country of interest.  
```{r plot-confirmed}
melted.dt <- melt(country_confirmed.dt, id.vars='Last_Update', variable.name = 'Cases', 
                  value.name='Number.Reported')
ggplot(melted.dt, aes(x=Last_Update, y=Number.Reported)) +
    geom_point() + 
    geom_line()
```

Here, we show the number of reported deaths and recoveries each day for our country
of interest.  Notice how different are the maximum values in the y-column s between
this plot and the first one.  This is why it was convenient to split them up this way,
otherwise these two plots would be very small and hard to read!
```{r plot-cases}
melted.dt <- melt(country_cases.dt, id.vars='Last_Update', variable.name = 'Cases', 
                  value.name='Number.Reported')
ggplot(melted.dt, aes(x=Last_Update, y=Number.Reported, color=Cases)) +
    geom_point() + 
    geom_line()
```

    

# Further reading
See the following links for information on the following topics:

* [*Data tables*](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) in R


* [*R for Data Science*](https://r4ds.had.co.nz/).  An excellent resource for those
who want to delve into other data sets on their own, learning data analysis from 
the ground up in R.  Accessible to all levels of experience!
